---
title: "Title Here"
author: 
  - Nya Gasowski1^[American University]
  - Rian Russell2^[American University]
  - Will Palmer3^[American University]
format: 
  pdf:
    number-sections: true
    toc: false
    geometry: 
      - top=3cm
date: 2026-01-09
date-format: iso
documentclass: article
editor: source
abstract: "In this paper, we examine the relationship between upzoning and the prevalence of non-violent crime, particularly auto theft and theft from auto in Minneapolis and St. Paul, Minnesota. In 2020, Minneapolis implemented the Minneapolis 2040 Plan, becoming the first U.S. city to eliminate single-family zoning, a reform expected to increase residential density. We consider this policy change as an experimental opportunity to study whether upzoning is in fact associated with real changes in auto-related property crime. Using neighborhood-level monthly crime data from 2017 to 2022, we construct a dataset combining police incident reports from Minneapolis and St. Paul. While only Minneapolis experienced the change in zoning, we consider St. Paul to be a control variable in this context due to its close proximity to Minneapolis. We found (add results from linear model here)."
---

# Title

# Introduction

In recent years, cities across the United States have increasingly turned to upzoning as a tool to address housing shortages, affordability concerns, and urban sprawl. Upzoning allows for higher-density development in previously low-density areas and has the potential to substantially reshape neighborhood composition and activity. While much literature examines the economic and housing-market effects of zoning reform, less is known about its implications for public safety. While many studies predict a positive relationship between density and crime, they often have considerably unique variables and limitations. Other research suggests that upzoning may instead reduce certain types of crime through natural surveillance, especially that which draw on Jane Jacobs’ “eyes on the street” theory. This is particularly in contrast to commercial density, which differs in that populations are often transient, creating gaps in surveillance while residential density offers more permanent awareness (Mitre-Becerril & MacDonald, 2024). When putting the Minneapolis 2040 plan into legislation, policy makers did so with the intended goal of improving housing affordability, transportation access, environmental sustainability, economic development, and public safety (City of Minneapolis DCPED, n.d.). The plan highlights pedestrian centered design, lighting, mixed-income housing, and crime prevention through environmental design, suggesting that increased residential density is intended to enhance neighborhood safety rather than exacerbate crime. We seek to explore how effective this legislation has really been in affecting crime prevalence, and if it reflects the intentions of the City of Minneapolis.

Our analysis contributes to an ongoing debate about whether upzoning exacerbates crime. Considering theories of social disorganization and the potential of increased population due to the upzoning policy (Braithwaite, 1975), we hypothesize that upzoning increases auto theft and theft from auto by increasing the number of potential targets and offenders within an area. We test this hypothesis using monthly, neighborhood-level crime data from 2017 to 2022. We estimate difference-in-differences models, including Poisson generalized linear models (appropriate for count data), to compare crime trends before and after the reform and between Minneapolis and St. Paul.

# Zoning, Density, and Crime

The influence of zoning density, which is the amount of development permitted in a certain unit of land, on crime rates has become an increasingly researched topic, drawing from a diverse range of fields, from criminology to urban planning. It is important to determine whether new zoning laws have led to an increase in crime so that these laws can be analyzed and amended if necessary.

Many studies focus on psychological and criminological theories in examining the effects of density on crime. Clement et al. (2025), who look at urban density growth in different counties in the US and its effect on crime, refer to the social disorganization and routine activities theories. Ma et al. (2025), who similarly look at urban density growth in different counties but in China, bring up Maslow’s hierarchy of needs and Becker’s economic model of crime, which views offenders as individuals who weigh the benefits and costs of their actions.

Recent research includes a variety of results. He and Li (2021), who look specifically at property crime in Dallas and Fort Worth, come to the conclusion that commercial and mixed-land use development, the number of transit facilities, and alcohol-related establishments like bars are positively associated with property crime rates in the two Texas cities, although their results are not generalizable. Clement et al. (2025) find that migration, which is the inflow of residents minus the outflow, is positively associated with vehicle theft rate, but many of their hypotheses were rejected. Ma et al. (2025) find that there is a positive correlation between urban density and urban crime rates, but their results are specific to China. Therefore, it is imperative that more research is done.

There are also a number of methods that quantitative research uses, including ordinary least squares with lagged first-differences (as criminological research recommends using a three-year lag for longitudinal models), geographically weighted negative binomial regression, and two-stage least squares estimates (He & Li, 2021; Clement et al., 2025; Ma et al, 2025). However, (a number of other studies use GLM)

In 2020, Minneapolis became the first city in the US to eliminate single-family zoning, which is defined as restricting residential areas to single-family homes, in its 2040 plan. The goal of single-family zoning is to maintain low density neighborhoods, meaning that this plan could lead to higher density neighborhoods (Gu & Murno, 2025).

# Data and Methods

Tba

```{r}
#| label: setup
#| echo: false
#| message: false

library(tidyverse)
library(lubridate)
library(marginaleffects)
```

```{r}
#| echo: false
#| message: false

MA2020 <- read_csv("https://raw.githubusercontent.com/will-palmer-6/FinalProj/refs/heads/main/Institute_Project/Data/Minneapolis/Police_Incidents_2020%20(1).csv")
MA2018_1 <- read_csv("https://raw.githubusercontent.com/will-palmer-6/FinalProj/refs/heads/main/Institute_Project/Data/Minneapolis/Police_Incidents_2018_435460759263840626.csv")
MA2018_2 <- read_csv("https://raw.githubusercontent.com/will-palmer-6/FinalProj/refs/heads/main/Institute_Project/Data/Minneapolis/Police_Incidents_2018_PIMS.csv")
MA2019 <- read_csv("https://raw.githubusercontent.com/will-palmer-6/FinalProj/refs/heads/main/Institute_Project/Data/Minneapolis/Police_Incidents_2019.csv")
MA2021 <- read_csv("https://raw.githubusercontent.com/will-palmer-6/FinalProj/refs/heads/main/Institute_Project/Data/Minneapolis/Police_Incidents_2021.csv")
MA2022 <- read_csv("https://raw.githubusercontent.com/will-palmer-6/FinalProj/refs/heads/main/Institute_Project/Data/Minneapolis/Police_Incidents_2022_-3717892713598614431.csv")
MA2017 <- read_csv("https://raw.githubusercontent.com/will-palmer-6/FinalProj/refs/heads/main/Institute_Project/Data/Minneapolis/Police_Incidents_2017_-5030148359353074920.csv")


# COMBINING MINNEAPOLIS DATA (2018-2022)
MA182_22 <- bind_rows(MA2020, MA2018_2, MA2019, MA2021, MA2022) 

# FILTERING AND RENAMING/REFORMATTING COLUMNS FOR FUTURE BINDING
MA182_22 <- MA182_22 |>
  select(publicaddress, precinct, reportedDate, offense, description, neighborhood, centerLat, centerLong) |>
  rename(c(lat = centerLat, long = centerLong, reporteddate = reportedDate))
MA182_22 <- MA182_22 |>
  mutate(date = as.Date(parse_date_time(reporteddate, orders = c("ymd HMSz", "mdy HMS p"))))
  
# COMBINING MINNEAPOLIS DATA (2017-2018)
MA17_181 <- bind_rows(MA2017, MA2018_1)

# FILTERING AND RENAMING/REFORMATTING COLUMNS FOR FUTURE BINDING
MA17_181 <- MA17_181 |>
  select(PublicAddress, Precinct, ReportedDate, Offense, Description, Neighborhood, Lat, Long) |>
  mutate(date = as.Date(parse_date_time(ReportedDate, orders = "mdy HMS p")))
names(MA17_181) <- tolower(names(MA17_181))

#BINDING ALL MINNEAPOLIS FROM 2017-2022 AND TRANSFORMING DATE TO MONTH
MA17_22 <- bind_rows(MA17_181, MA182_22)


# READING IN ST. PAUL DATA, CLEANED ON REDIVIS
STP <- read_csv("https://raw.githubusercontent.com/will-palmer-6/FinalProj/refs/heads/main/Institute_Project/Data/renaming_output.csv")


# COMBINING ST PAUL AND MINNEAPOLIS, CONDENSING INTO CRIME COUNT OVER DESIRE YEARS, SETTING DAYS TO MONTHS
MN_Crime <- bind_rows(MA17_22, STP) |>
  select(neighborhood, date, offense, description) |>
  mutate(month = floor_date(date, "month"))

MN_Crime_Grouped <- MN_Crime |>
  group_by(month, neighborhood) |>
  summarize(crime = n()) |>
  filter(month >= as.Date("2017-01-01")) |>
  filter(month <= as.Date("2022-12-31")) 

# BALANCING TABLE FOR MONTH NEIGHBORHOODS WITH NO CRIME DATA
MN_Crime_Grouped <- MN_Crime_Grouped |>
  ungroup() |>
  complete(neighborhood, month, fill = list(crime = 0)) 

# CREATING TREATMENT AND POST JAN 1 2020 VARIABLES FOR LINEAR MODEL
MN_Crime_Grouped <- MN_Crime_Grouped |>
  mutate(treatment = case_when(neighborhood == tolower(neighborhood) ~ 0, is.na(neighborhood) ~ 1, TRUE ~ 1),
         post = ifelse(month < as.Date("2020-01-01"), 1, 0))
```

# \[Our Results Section Title Here\]

Tba

# Discussion

Tba

\clearpage

# References

